---
title: "4_PCA_plotting_code"
author: "Laura Tibbs"
date: "November 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goal:
Plot PC output from Eigensoft to approximately reproduce Fig 1-2 from Blanca, et al. 2015.

Set working directory to the location of this `.Rmd` file:
```{r}
# This needs to be changed to reflect the location of the GitHub repository and this code within it on any given computer that runs the code. 
# In my case, this is:
# setwd("C:/Users/ltibbs/Box Sync/ISU/2018 fall/BCB 546X Computational Skills for Bio Data/BCB546X_MadMaize_final/code/PCA")
```

Load libraries:
```{r}
library(tidyverse)
```

## Fig 1:
Reproduce Fig 1: PCA using all uniquely named accessions.

The Eigensoft package produced a plot (`Fig1_Eigensoft_ploteig.pdf`) that is very similar to Fig. 1 in the paper, but it differs in that the percentage of variability explained by each PC is shown in Fig 1 but not in the output from Eigensoft; also, PC 1 is on the x axis and PC 2 is on the y axis in Eigensoft's figure, but this is reversed in Fig 1. So, I will graph the data from Eigensoft using R to make these changes.

Load data and tidy it:
```{r}
# read data
fig1 <- read.table("eigensoft/full.eigensoft.pca.evec")

# 
```

## Fig 2:
for plotting circles around stuff--look at https://stackoverflow.com/questions/13577918/plotting-a-curve-around-a-set-of-points